Computational biology in the 21st century: Algorithms that scale

### 2015-07-29 main.tex
more cutting, and fix quotes

As mentioned earlier, function follows form, so in addition to
sequence and expression, structure plays an important role in biological data science.
Protein structure data, primarily determined through X-ray crystallography or
nuclear magnetic resonance (NMR) spectroscopy, describes the coordinates of 
each atom, in Angstroms, of a protein in three-dimensional space.
This information is important to understanding how a protein functions, because
proteins interact physically with other molecules and are geometrically constrained.
Protein structure data are commonly stored in the Protein Data Bank (PDB); 
each PDB entry has an associated resolution indicating the accuracy of the
structural description.
Thus a PDB entry at its most basic level contains a list of atoms, along with
which amino acid they belong to and their spatial coordinates as floating-point
numbers.

Chemical compounds represent another source of structural data.
Although generically much smaller than proteins, their structures are more general
as they are not constrained in composition to amino acids.
Determined by X-ray crystallography, NMR, electron microscopy and other 
techniques, chemical compounds---many of which are useful pharmaceuticals---can be
represented as labeled graphs (or equivalently, adjacency matrices).
In such graphs, the vertices are atoms (labeled with a chemical element) and
the edges are bonds between atoms (labeled with a bond type, such as single or
double).
Chemical compounds are stored in databases such as PubChem, and are often searched
for by medicinal chemists as potential agonists for protein targets.

@@

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
obesity~\cite{greenblum2012metagenomic}.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
The difficulty is that the datasets required to shine any light on these 
questions, namely from `shotgun' metagenomics, are gigantic and vastly more 
complex than the 16S metagenomic studies that merely classify the bacteria 
based on 
their divergence with respect to a specific gene (namely, the 16S ribosomal 
subunit).
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

@@

While gene expression data lends itself to cluster analysis and probabilistic 
approaches, the high dimensionality and noise in the data presents significant 
challenges~\cite{nguyen2002tumor}.
Principal Component Analysis has shown promise in reducing the dimensionality
of gene expression 
data~\cite{raychaudhuri2000principal,alter2000singular,yeung2001principal,misra2002interactive}.
Such data and its challenges have been the focus of other 
articles~\cite{berger2013computational}, and thus will be only lightly touched 
upon here.

@@

Techniques for reference-based read mapping typically rely on algorithmic 
approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
which provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
based on the BWT, which provides efficient storage as well as fast search.
BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
Bowtie2~\cite{langmead2012fast} mapper further relies on the FM-index for 
efficient mapping of NGS reads.
The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
coupled with dynamic programming in a compressed representation of the 
reference genome, in order to prune the search space 
when mapping reads to a reference genome.
Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2014mrsfast} use an `approximate seed' approach to indexing
the space of possible matches, likewise pruning the search space; however, the
bulk of its runtime is spent on the extend phase.
State-of-the-art mapper mrsFAST-Ultra achieves improvements in efficiency based on machine architecture rather than leveraging redundancy in the data itself with near-perfect sensitivity, but
only for the case where there are no insertions and deletions (indels) \cite{hach2014mrsfast}.
Even with these approaches, read mapping remains a significant bottleneck in
genomic research~\cite{hach2010mrsfast}.

@@

Recent work on gene expession has explored additional ways to exploit the high-dimensional 
structure of the data.
SPARCLE (SPArse ReCovery of Linear combinations of Expression)~\cite{prat2011recovering}
brings ideas from compressed sensing~\cite{candes2005decoding,candes2006stable,donoho2006compressed} 
to gene expression analysis.
Specifically, SPARCLE relies on the idea that a randomly-chosen set of points 
are unlikely to be close in a lower-dimensional linear subspace.
Thus, SPARCLE approximates a gene's expression profile as a linear combination
of a few other genes' profiles, and then applies supervised learning to
predict relationships between genes.
Another recent and novel approach to exploiting the structure of gene expression
space is Parti (Pareto task inference)~\cite{hart2015inferring}, which describes a set of
data as a polytope, and infers the specific tasks represented by vertices of
that polytope from the features most highly enriched at those vertices.
In human breast tumors, and in mouse tissues, the expression data were
well-described by tetrahedra whose vertices were enriched for different tumor
types and biological functions, from which four distinct tasks are 
inferred~\cite{hart2015inferring}.

@@

Beyond redundancy, however, another attribute of large biological data sets
stands out.
Any high-dimensional data set, such as amino acid sequences of a bounded 
length, can be described as inhabiting a vector space.
For example, given an alphabet of 20 amino acids and a length bounded at 
500\footnote{While the largest-known protein, titin, is over 33,000 amino acids 
in length, proteins are typically described as comprising \emph{domains} that 
evolve and fold independently; domains rarely exceed 500 amino acids.}, one 
could
enumerate $20^{500} \approx 3 \times 10^{650}$ distinct protein sequences, as 
of July 
2015, the NCBI's exhaustive ``NR'' database lists $6.9 \times 10^{7}$ sequences.
This surprisingly lower number is not simply due to undersampling; thanks to 
evolution, only those protein
sequences which exhibit useful biological function should survive, and most
random sequences of amino acids would not be expected to form stable structures.
Thus, not only do biological data exhibit redundancy, they also tend not to 
inhabit anywhere near the entire feasible space (Figure~\ref{fig:dataspace}).
It seems that physical laws have constrained the data to a particular subspace
of the enumerable Cartesian space.

@@

One key insight related to redundancy
%as to why the compressive omics approach allows computation 
%that scales sublinearly in the size of the data 
is that such data sets exhibit
low \emph{metric entropy}~\cite{yu2015entropy}.
That is, for a given cluster radius $r_c$ and a database $D$, the number $k$ 
of clusters needed to cover $D$ is bounded by $N_{r_c} (D)$, the metric 
entropy, which is relatively small compared to
$|D|$, the number of entries in the database (Figure~\ref{fig:dataspace}).
In contrast, if the points were uniformly distributed about the Cartesian space,
$N_{r_c} (D)$ would be larger.

@@

Cluster-based search, as exemplified by ``compressive omics,'' can perform
approximate search within a radius $r$ of a query $q$ on a database $D$ with 
fractal dimension $d$ and metric entropy $k$ at the scale $r_c$ in time
proportional to
\begin{displaymath}
    O\Bigg(
    \underbrace{k}_{\textrm{metric entropy}} +
    \overbrace{\left|B_D(q,r)\right|}^{\textrm{output size}}
    \underbrace{\left(\frac{r+2r_c}{r}\right)^d}_{\textrm{scaling factor}}
     \Bigg) .
\end{displaymath}

>>>>

Chemical compounds represent a source of structural data.
Determined by X-ray crystallography, NMR, electron microscopy and other 
techniques, chemical compounds---many of which are useful pharmaceuticals---can be
represented as labeled graphs (or equivalently, adjacency matrices).
In such graphs, the vertices are atoms (labeled with a chemical element) and
the edges are bonds between atoms (labeled with a bond type, such as single or
double).
Chemical compounds are stored in databases such as PubChem, and are often searched
for by medicinal chemists as potential agonists for protein targets.

@@

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment}, and 
obesity~\cite{greenblum2012metagenomic}.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
The difficulty is that the datasets required to shine any light on these 
questions, namely from `shotgun' metagenomics, are gigantic and vastly more 
complex than the 16S metagenomic studies that merely classify the bacteria 
based on 
their divergence with respect to a specific gene (namely, the 16S ribosomal 
subunit).
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

@@

While gene expression data lends itself to cluster analysis and probabilistic 
approaches, the high dimensionality and noise in the data presents significant 
challenges~\cite{nguyen2002tumor}.
Principal Component Analysis has shown promise in reducing the dimensionality
of gene expression 
data~\cite{raychaudhuri2000principal,alter2000singular}.
Such data and its challenges have been the focus of other 
articles~\cite{berger2013computational}, and thus will be only lightly touched 
upon here.

@@

Techniques for reference-based read mapping typically rely on algorithmic 
approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
which provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
based on the BWT, which provides efficient storage as well as fast search.
BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
Bowtie2~\cite{langmead2012fast} mapper further relies on the FM-index for 
efficient mapping of NGS reads.
The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
coupled with dynamic programming in a compressed representation of the 
reference genome, in order to prune the search space 
when mapping reads to a reference genome.
Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2010mrsfast} use an `approximate seed' approach to indexing
the space of possible matches, likewise pruning the search space; however, the
bulk of its runtime is spent on the extend phase.
State-of-the-art mapper mrsFAST-Ultra achieves improvements in efficiency based on machine architecture rather than leveraging redundancy in the data itself with near-perfect sensitivity, but
only for the case where there are no insertions and deletions (indels) \cite{hach2014mrsfast}.
Even with these approaches, read mapping remains a significant bottleneck in
genomic research~\cite{hach2010mrsfast}.

@@

Recent work on gene expession has explored additional ways to exploit the high-dimensional 
structure of the data.
SPARCLE (SPArse ReCovery of Linear combinations of Expression)~\cite{prat2011recovering}
brings ideas from compressed sensing~\cite{candes2005decoding} 
to gene expression analysis.
Specifically, SPARCLE relies on the idea that a randomly-chosen set of points 
are unlikely to be close in a lower-dimensional linear subspace.
Thus, SPARCLE approximates a gene's expression profile as a linear combination
of a few other genes' profiles, and then applies supervised learning to
predict relationships between genes.
Another recent and novel approach to exploiting the structure of gene expression
space is Parti (Pareto task inference)~\cite{hart2015inferring}, which describes a set of
data as a polytope, and infers the specific tasks represented by vertices of
that polytope from the features most highly enriched at those vertices.
In human breast tumors, and in mouse tissues, the expression data were
well-described by tetrahedra whose vertices were enriched for different tumor
types and biological functions, from which four distinct tasks are 
inferred~\cite{hart2015inferring}.

@@

Beyond redundancy, however, another attribute of large biological data sets
stands out.
Any high-dimensional data set, such as amino acid sequences of a bounded 
length, can be described as inhabiting a vector space.
For example, given an alphabet of 20 amino acids and a length bounded at 
500\footnote{While the largest-known protein, titin, is over 33,000 amino acids 
in length, proteins are typically described as comprising \emph{domains} that 
evolve and fold independently; domains rarely exceed 500 amino acids.}, one 
could
enumerate $20^{500} \approx 3 \times 10^{650}$ distinct protein sequences, as 
of July 
2015, the NCBI's exhaustive `NR' database lists $6.9 \times 10^{7}$ sequences.
This surprisingly lower number is not simply due to undersampling; thanks to 
evolution, only those protein
sequences which exhibit useful biological function should survive, and most
random sequences of amino acids would not be expected to form stable structures.
Thus, not only do biological data exhibit redundancy, they also tend not to 
inhabit anywhere near the entire feasible space (Figure~\ref{fig:dataspace}).
It seems that physical laws have constrained the data to a particular subspace
of the enumerable Cartesian space.

@@

One key insight related to redundancy
is that such data sets exhibit
low \emph{metric entropy}~\cite{yu2015entropy}.
That is, for a given cluster radius $r_c$ and a database $D$, the number $k$ 
of clusters needed to cover $D$ is bounded by $N_{r_c} (D)$, the metric 
entropy, which is relatively small compared to
$|D|$, the number of entries in the database (Figure~\ref{fig:dataspace}).
In contrast, if the points were uniformly distributed about the Cartesian space,
$N_{r_c} (D)$ would be larger.

@@

Cluster-based search, as exemplified by `compressive omics,' can perform
approximate search within a radius $r$ of a query $q$ on a database $D$ with 
fractal dimension $d$ and metric entropy $k$ at the scale $r_c$ in time
proportional to
\begin{displaymath}
    O\Bigg(
    \underbrace{k}_{\textrm{metric entropy}} +
    \overbrace{\left|B_D(q,r)\right|}^{\textrm{output size}}
    \underbrace{\left(\frac{r+2r_c}{r}\right)^d}_{\textrm{scaling factor}}
     \Bigg) .
\end{displaymath}

### 2015-07-29 supplement.tex
more cutting, and fix quotes

>>>>

As mentioned earlier, function follows form, so in addition to
sequence and expression, structure plays an important role in biological data science.
Protein structure data, primarily determined through X-ray crystallography or
nuclear magnetic resonance (NMR) spectroscopy, describes the coordinates of 
each atom, in Angstroms, of a protein in three-dimensional space.
This information is important to understanding how a protein functions, because
proteins interact physically with other molecules and are geometrically constrained.
Protein structure data are commonly stored in the Protein Data Bank (PDB); 
each PDB entry has an associated resolution indicating the accuracy of the
structural description.
Thus a PDB entry at its most basic level contains a list of atoms, along with
which amino acid they belong to and their spatial coordinates as floating-point
numbers.
